---
title: "Exclude articles not relevant to EHR-based automated phenotyping"
author: "Siyue Yang, Jessica Gronsbell"
date: "05/18/2022"
output: pdf_document
knit: (function(inputFile, encoding) {
      out_dir <- "../report";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(kableExtra)
```

```{r}
df <- read.csv("../data/annotations050822.csv", skip = 1)
```

After merged articles of PubMed and Web of Sciences by `1-merging.rmd`, there are 852 papers.  

```{r}
init <- df %>% 
  group_by(Source) %>%
  summarise(n = n()) 

init <- init %>%
  add_row(Source = "Total", n = sum(init$n)) 

init %>%
  kbl(booktabs = T, caption = "Number of papers included after merging papers from two database") %>%
  kable_paper("striped") %>%
  kable_styling(position = "center", latex_options = "HOLD_position")
```

We annotated based on the `annotations050822.csv` file. Our analysis consists of 3 steps:

- **Screen for title and abstract** and exclude the papers by our inclusion or exclusion criteria. 
- **Screen for full-text** and exclude the papers by the same inclusion and exclusion criteria. 
- **Include relevant papers** that are cited by papers in our pool or we encountered in our previous work.

```{r}
## Step 1: Title and abstract review. 
step1 <- df %>%
  group_by(Primary_reason_for_exclusion_TA_SY) %>%
  summarise(n = n()) %>%
  filter(Primary_reason_for_exclusion_TA_SY != "")

colnames(step1) <- c("Primary reason for exclusion", "N") 
```

```{r}
## Step 2: Full-text review. 
step2 <- df %>%
  filter(Include_FR_SY == 0) %>%
  group_by(Primary_reason_for_exclusion_FR_SY) %>%
  summarise(n = n()) 

colnames(step2) <- c("Primary reason for exclusion", "N") 
```

```{r}
exclusion <- merge(step1, step2, by = "Primary reason for exclusion", all = TRUE)
colnames(exclusion) <- c("Primary reason for exclusion", "Title and abstract review", "Full-text review")

# Replace NA with 0.
exclusion <- exclusion %>%
  mutate_all(funs(replace_na(., 0))) %>%
  mutate(Total = `Title and abstract review` + `Full-text review`) 

exclusion %>%
  add_row(`Primary reason for exclusion` = "Total", 
          `Title and abstract review` = sum(exclusion$`Title and abstract review`), 
          `Full-text review` = sum(exclusion$`Full-text review`),
          `Total` = sum(exclusion$Total)) %>%
  kbl(booktabs = T, caption = "Number of papers excluded") %>%
  kable_paper("striped") %>%
  kable_styling(position = "center", latex_options = "HOLD_position")
```

We also report the number of papers removed in Table 3. 

```{r}
before_screnning <- init %>%
  filter(Source == "Total") %>%
  select(n) %>%
  pull()

excluded_title <- sum(exclusion$`Title and abstract review`)

excluded_full_text <- sum(exclusion$`Full-text review`)

process <- c("Before screening", "Title and abstract review", "Full-text review")
removed <- c(0, excluded_title, excluded_full_text)
added <- c(0, 0, 0)
remained <- c(before_screnning, 
              before_screnning - excluded_title, 
              before_screnning - excluded_title - excluded_full_text) 

fig_1 <- data.frame(process, removed, added, remained)

colnames(fig_1) <- c("Process", "Number of papers removed", "Number of papers added", "Number of paeprs remained")

fig_1 %>%
  kbl(booktabs = T, caption = "Number of paper removed and added") %>%
  kable_paper("striped") %>%
  kable_styling(position = "center", latex_options = c("scale_down", "HOLD_position"))
```
