---
title: "A summary of EHR-based phenotyping article annotation"
author: "Siyue Yang, Jessica Gronsbell"
date: "05/25/2022"
output: 
  pdf_document:
    toc: true
    number_sections: true
knit: (function(inputFile, encoding) {
      out_dir <- "../report";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, eval = T}
# Load necessary packages
library(ggplot2)
library(tidyverse)
library(kableExtra)
library(ggsci)
library(grid)
library(gridExtra)
library(ggvenn)
library(ggpattern)
library(cowplot)

# Functions for analysis
source('analysis_functions.R')
```

```{r, eval = T}
# Read in annotation file
df_all <- read.csv("../data/annotations050822.csv", skip = 1)
df <- df_all %>% filter(Include_FR_SY == 1)

# Create variable indicating use of unstructured data
df$Unstructured <- (df$Unstructured_data_type != "")

# Create variable indicating use of structured data
df$Structured <- (df$Structured_data_type != "")

# Create variable indicating types of data
df$Data_type[df$Structured == 1] <- "Structured"
df$Data_type[df$Unstructured == 1] <- "Free-text"
df$Data_type[df$Unstructured + df$Structured == 2] <- "Both structured and free-text"
# As the paper used non-language unstructured data and structured data,
# we include it as structured data. 
df$Data_type[df$Unstructured_data_language == "Non-language"] <- "Structured"

# Create variable indicating use of traditional ML method
df$Traditional[df$DL == ''] <- "Traditional machine learning"
df$Traditional[df$DL != ''] <- "Deep learning"
```

\clearpage
# Overview 

```{r}
df$ML_type <- factor(df$ML_type, levels = c("Supervised", 
                                             "Semi-supervised", 
                                             "Weakly-supervised", 
                                             "Unsupervised"))
stacked_bar(df, "ML_type", "Traditional", order_by_count = FALSE)
```

```{r, eval = F}
# Save in eps.
g1 <- stacked_bar(df, "ML_type", "Traditional", order_by_count = FALSE, large_fig = TRUE)
ggsave(g1, file="../figure/fig2.eps", width=40, height=20, units="cm", dpi=600) 
```

## Traditional ML method 

```{r}
# Traditional ML method.
method_unnested <- unnest_string_var(df, "Traditional_ML_method")
print_tables(method_unnested, "ML_type", "Traditional_ML_method_unnested", top_count = 20, restric_count = TRUE,
             title = "Common traditional machine learning methods (Count > 1)")
print_multiple_items(method_unnested, "traditional machine learning methods")
```

## DL method

```{r}
# DL method. 
method_unnested <- unnest_string_var(df, "DL_method")
print_tables(method_unnested, "DL_method_unnested", "ML_type", top_count = 10,
             title = "Deep supervised learning methods", restric_count = TRUE)
print_multiple_items(method_unnested, "deep learning methods")
```


```{r}
traditional_supervised <- df %>% filter(ML_type == "Supervised") %>% filter(DL_method == '')
deep_supervised <- df %>% filter(ML_type == "Supervised") %>% filter(DL_method != '')
semi_supervised <- df %>% filter(ML_type == "Semi-supervised") 
weakly_supervised <- df %>% filter(ML_type == "Weakly-supervised") 
un_supervised <- df %>% filter(ML_type == "Unsupervised") 

traditional_supervised$Category <- "Traditional supervised"
deep_supervised$Category <- "Deep supervised"
semi_supervised$Category <- "Semi-supervised"
weakly_supervised$Category <- "Weakly-supervised"
un_supervised$Category <- "Unsupervised"
```

### Deep neural network variants

```{r, fig.height=6}
method_unnested <- unnest_string_var(deep_supervised, "RNN_subname")
p1 <- horizontal_bar(method_unnested, "RNN_subname_unnested", title = "RNN", ylim = 10)

method_unnested <- unnest_string_var(deep_supervised, "CNN_subname")
p2 <- horizontal_bar(method_unnested, "CNN_subname_unnested", title = "CNN", ylim = 12)

method_unnested <- unnest_string_var(deep_supervised, "BERT_subname")
p3 <- horizontal_bar(method_unnested, "BERT_subname_unnested", title = "BERT", ylim = 10)

method_unnested <- unnest_string_var(deep_supervised, "FFNN_subname")
p4 <- horizontal_bar(method_unnested, "FFNN_subname_unnested", title = "FFNN", ylim = 12)

plot_grid(p1, p2, p3, p4, ncol = 2, nrow = 2, align = "v", axis = "l")
```

\clearpage 
# Phenotype

```{r}
# Unique phenotypes
pheno_unnested <- unnest_string_var(df %>% filter(Competition_data_name == ""), "Phenotype")
length(unique(pheno_unnested$Phenotype_unnested))

# more than 1 phenotype
sum(df$Total_number_of_phenos > 1, na.rm = TRUE)

# binary phenotype
sum(df$Pheno_binary > 0, na.rm = TRUE)

sum(df$Pheno_severity > 0, na.rm = TRUE)

sum(df$Pheno_temporal > 0, na.rm = TRUE)
```

```{r, fig.height = 8}
# Phenotype - non-competition
plot_all_top_pheno(traditional_supervised,
                               deep_supervised,
                               semi_supervised,
                               weakly_supervised,
                               un_supervised)
```

```{r}
# Save figures in eps format and set large figure to TRUE.
# Phenotype - non-competition
g1 <- plot_all_top_pheno(traditional_supervised,
                               deep_supervised,
                               semi_supervised,
                               weakly_supervised,
                               un_supervised)
ggsave(g1, file="../figure/fig3.eps", width=35, height=40, units="cm", dpi=600) 
```

## More nuanced phenotype

```{r, eval = F}
# All phenotype with count > 1.
temporal_unnested <- unnest_string_var(df, "Phenotype")
horizontal_bar(temporal_unnested, "Phenotype_unnested", 
               restrict_count = TRUE,
               ylim = 10,
               legend_cap = "Temporal phenotype category")
```



\clearpage
# Data source

## Summary

```{r}
data_source <- rbind(traditional_supervised, deep_supervised, 
                     semi_supervised, weakly_supervised, un_supervised)
data_source$Category <- factor(data_source$Category, levels = c("Traditional supervised", 
                                                                "Deep supervised", 
                                                                "Semi-supervised", 
                                                                "Weakly-supervised", 
                                                                "Unsupervised"))
data_source$Data_type <- factor(data_source$Data_type, levels = c("Structured", 
                                                                  "Free-text",
                                                                  "Both structured and free-text", "Image"))
stacked_bar(data_source, "Category", "Data_type", order_by_count = FALSE) +
  scale_y_continuous(n.breaks = 10)
```

```{r}
# Save figure in eps. 
g2 <- stacked_bar(data_source, "Category", "Data_type", order_by_count = FALSE, large_fig = TRUE) +
  scale_y_continuous(n.breaks = 10)
ggsave(g2, file="../figure/fig4.eps", width=40, height=20, units="cm", dpi=600) 
```

```{r}
# Get data source counts for paper

```
```{r}
print_summary_table(list(traditional_supervised, deep_supervised, semi_supervised,
                         weakly_supervised, un_supervised, df)) 
```

TSL = Traditional supervised learning.  DSL = Deep supervised learning.  DRL = Reinforcement deep learning.  SSL = Semi-supervised learning.  WSL = Weakly-supervised learning.  US = Unsupervised learning. 

## Structured and unstructured data type

```{r, eval = F}
data_unnested <- unnest_string_var(df, "Data_type")
print_tables(data_unnested, "Data_type", top_count = 10, title = "Types of data used.")
```

```{r, fig.height=8}
data_unnested <- unnest_string_var(df %>% filter(Structured == 1), "Structured_data_type")


data_unnested <- data_unnested %>%
  mutate(Structured_data_type_unnested = case_when(Structured_data_type_unnested == "Diagnosis" ~ "Diagnoses", TRUE ~ Structured_data_type_unnested)) 

p1 <- horizontal_bar(data_unnested, "Structured_data_type_unnested", 
                     string_wrap = 20,restrict_count = TRUE, 
                     title = "Structured data type", ylim = 60, label_text = TRUE,
                     xlab = "Number of articles")

print_multiple_items(data_unnested, "structured data type")

data_unnested <- unnest_string_var(df %>% filter(Unstructured == 1), "Unstructured_data_type")

data_unnested <- data_unnested %>%
  mutate(Unstructured_data_type_unnested = case_when(Unstructured_data_type_unnested == "Clinical notes" ~ "Unspecified", TRUE ~ Unstructured_data_type_unnested))
p2 <- horizontal_bar(data_unnested, "Unstructured_data_type_unnested", 
                     string_wrap = 20, restrict_count = TRUE, 
                     title = "Clinical note type", ylim = 60, label_text = TRUE,
                     xlab = "Number of articles")
print_multiple_items(data_unnested, "unstructured data type")

plot_grid(p1, p2, ncol = 1, align = "v", axis = "l") 
```

\clearpage

## Institutions

```{r}
count_unnested <- unnest_string_var(df, "Country")
print_tables(count_unnested, "Country_unnested", 
             top_count = 10, restric_count = FALSE, hold_pos = TRUE)
```



## Openly-available data

```{r}
method_unnested <- unnest_string_var(df, "Competition_data_name")
print_multiple_items(method_unnested, "Competition data")

print_tables(method_unnested, "Competition_data_name_unnested", "ML_type", "Traditional", 
             top_count = 10, restric_count = FALSE, hold_pos = TRUE)
  
```

```{r, eval = F}
stacked_bar(method_unnested %>% filter(Competition_data_name != ""), 
            "Competition_data_name", "Traditional", string_wrap = 20)
```


```{r}
df %>% filter(Competition_data_name != "") %>% nrow()
```


```{r}
method_unnested <- unnest_string_var(df %>% filter(Competition_data_name == ""), "Data_source")
open_data <- c("MIMIC-III database", "MTSamples database")
method_unnested <- method_unnested %>% filter(Data_source_unnested %in% open_data)

print_tables(method_unnested, "Data_source_unnested", "ML_type", "Traditional", top_count = 10, restric_count = FALSE)
```

\clearpage
# Terminology

```{r}
term_unnested <- unnest_string_var(df, "Terminology")

term_unnested <- term_unnested %>% 
  mutate(Terminology_unnested = case_when(Terminology_unnested == "ICD-9/10" ~ "ICD-9; ICD-10",
                                          TRUE ~ Terminology_unnested))

term_unnested <- unnest_string_var(term_unnested, "Terminology_unnested")

print_multiple_items(term_unnested, "terminologies")
print_tables(term_unnested, "Terminology_unnested_unnested", "ML_type", "Traditional", top_count = 20, restric_count = TRUE)
```



\clearpage
# NLP software

```{r}
method_unnested <- unnest_string_var(df, "NLP_software")
print_multiple_items(method_unnested, "NLP software")
print_tables(method_unnested, "NLP_software_unnested", "ML_type", "Traditional", top_count = 10, restric_count = TRUE)
```

# Emebddings

Embeddings were only used in deep supervised articles. 

```{r}
method_unnested <- unnest_string_var(deep_supervised, "Embedding_training_data")
print_tables(method_unnested, "Embedding_training_data_unnested", top_count = 10)
print_multiple_items(method_unnested, "embedding training data")

method_unnested <- unnest_string_var(deep_supervised, "Embedding")
print_tables(method_unnested, "Embedding_unnested", top_count = 10)
print_multiple_items(method_unnested, "embedding training methods")
```

# Validation and comparison

## Traditonal supervised ML vs. rule-based

```{r, fig.height=8}
plot_validate_metrics(traditional_supervised, comparator = "rule", large_fig = FALSE)
```


```{r, eval = F, results='hide'}
g4 <- plot_validate_metrics(traditional_supervised, comparator = "rule", large_fig = TRUE)
ggsave(g4, file="../figure/fig5.eps", width=30, height=30, units="cm", dpi=600) 
```

## Deep supervised ML vs. traditional supervised ML

```{r, fig.height = 8}
plot_validate_metrics(deep_supervised, comparator = "deep", large_fig = FALSE)
```

```{r, eval = F, results='hide'}
g6 <- plot_validate_metrics(deep_supervised, comparator = "deep", large_fig = TRUE)
ggsave(g6, file="../figure/fig6.eps", width=30, height=35, units="cm", dpi=600) 
```


## Weakly-supervised ML vs. rule-based algorithms

```{r, fig.height = 8}
plot_validate_metrics(weakly_supervised, comparator = "weakly", large_fig = FALSE)
```

## Weakly-supervised ML vs. traditional supervised ML

```{r, fig.height=8}
plot_validate_metrics(weakly_supervised, comparator = "weakly vs supervised", large_fig = FALSE)
```


# Model performance metric reporting


```{r}
method_unnested <- unnest_string_var(df %>% filter(ML_type != "Unsupervised"), "Model_performance_metrics")
print_tables(method_unnested, "Model_performance_metrics_unnested", "ML_type", "Traditional", 
             top_count = 20, restric_count = TRUE, col_width = 4.6)
```

```{r, eval= FALSE}
# Report at least one of the precision, recall, F-score
method_unnested %>%
  filter(Model_performance_metrics_unnested %in% c("Precision", "Recall", "F-score")) %>%
  group_by(PMID) %>%
  summarise(n = n()) %>%
  nrow()

# Report at least one of the precision, recall, F-score
method_unnested %>%
  filter(Model_performance_metrics_unnested %in% c("Brier score", "Calibration plots")) %>%
  group_by(PMID) %>%
  summarise(n = n()) 
```
